# Appendix 2: Meta-analysis of the association between age at menarche and depressive symptoms
  in early-mid adolescence
```{r include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(eval = FALSE)
```


## Overview
To determine the smallest effect size of interest (SESOI) for hypothesis 1a, we performed a meta-analysis of the association between age at menarche and depressive symptoms in early-to-mid adolescence. After screening all studies included in the meta-analysis by Ullsperger & Nikolas (2017), we included 5 studies of adolescents where this association was reported with sufficient detail:

- Joinson, C., Heron, J., Araya, R. & Lewis, G. Early menarche and depressive symptoms from adolescence to young adulthood in a UK cohort. *J. Am. Acad. Child Adolesc. Psychiatry* **52**, 591–598 (2013)
- Black, S. R. & Klein, D. N. Early menarcheal age and risk for later depressive symptomatology: the role of childhood depressive symptoms. *J. Youth Adolesc.* **41**, 1142–1150 (2012).
- Stice, E., Presnell, K. & Bearman, S. K. Relation of early menarche to depression, eating disorders, substance abuse, and comorbid psychopathology among adolescent girls. *Dev. Psychol.* **37**, 608–619 (2001).
- Lien, L., Haavet, O. R. & Dalgard, F. Do mental health and behavioural problems of early menarche persist into late adolescence? A three year follow-up study among adolescent girls in Oslo, Norway. *Soc. Sci. Med.* **71**, 529–533 (2010).
- Kaltiala-Heino, R., Marttunen, M., Rantanen, P. & Rimpelä, M. Early puberty is associated with mental health problems in middle adolescence. *Soc. Sci. Med.* **57**, 1055–1064 (2003).

The code for the meta-analysis is below. Our approach was inspired by [Baldwin et al.](https://springernature.figshare.com/articles/journal_contribution/Adverse_childhood_experiences_and_mental_health_A_genetically_informed_study_Registered_Report_Stage_1_Protocol_/13580777)

## Load libraries
```{r, message=FALSE}
library(effectsize) # for effect size conversions
library(compute.es) # for effect size conversions
library(metafor) # for meta-analysis
```

## Extract data from studies

## Joinson et al. (2013)
Estimates were taken from Table 2, early menarche compared to late menarche at age 14 (whole sample, unadjusted). 
```{r }
OR <- 1.94
OR_low_CI <- 1.39
OR_up_CI <- 2.71
log_OR <- log(OR) # convert odds ratios to log odds
log_SE <- ( log(OR_up_CI) - log(OR_low_CI) ) / 3.92 # convert odds ratio SE to log odds SE
var <- log_SE^2 # derive sampling variance 
id <- "joinson"
N1 <- 449 # depression at age 14 based on table 1
N2 <- 2843 - N1 # no depression

# transform log odds ratios to standardized beta/correlation coefficients
conversion <- lores(log_OR, var, N1, N2, dig=20, verbose=FALSE)
B <- conversion$r
B_var <- conversion$var.r

# put extracted data into dataframe
joinson <- data.frame(B, B_var, id)
joinson
```

## Black & Klein (2012)
Estimates were taken from Table 4, age at menarche onset predicting postpubertal depressive symptoms at age 13 (adjusted, since unadjusted was not reported). 
```{r}
b <- 0.37
b_se <- 0.09
B <- 0.14
B_se <- B*(b_se/b) # derive SE
B_var <- B_se^2 # derive sampling variance
id <- "black_klein"

# put extracted data into dataframe
black_klein <- data.frame(B, B_var, id)
black_klein
```

## Stice et al. (2001)
Estimates were taken from Table 3, early menarche compared to nonearly menarche, at age 13 (unadjusted). 
```{r}
B <- 0.14
b <- 1.61
b_low_ci <- 0.61
b_up_ci <- 2.62
B_low_ci <- B*(b_low_ci/b)
B_up_ci <- B*(b_up_ci/b)
B_var <- ( (B_up_ci - B_low_ci) / 3.92 )^2 # derive sampling variance
id <- "stice"

# put extracted data into dataframe
stice <- data.frame(B, B_var, id)
stice
```

## Lien et al. (2010)
Estimates were taken from Table 4, time of menarche below 11 years, assessed at age 15 (unadjusted).
```{r}
OR <- 2.5
OR_low_CI <- 1.6
OR_up_CI <- 3.9
log_OR <- log(OR) # convert odds ratios to log odds
log_SE <- ( log(OR_up_CI) - log(OR_low_CI) ) / 3.92 # convert odds ratio SE to log odds SE
var <- log_SE^2 # derive sampling variance
id <- "lien"
N1 <- 353 # depression at age 15 based on table 1
N2 <- 1013 # no depression at age 15 based on table 1

# transform log odds ratios to standardized beta/correlation coefficients
conversion <- lores(log_OR, var, N1, N2, dig=20, verbose=FALSE)
B <- conversion$r
B_var <- conversion$var.r

# put extracted data into dataframe
lien <- data.frame(B, B_var, id)
lien
```

## Kaltiala-Heino et al. (2003)
Early age at menarche (AAM) was defined by authors as 11 years or lower. They reported results for those with AAM 10 years or lower and 11 years separately. Therefore, we combined the 10 years or lower and 11 years estimates. Estimates were derived based on values reported in Table 1 (frequency distribution of adolescents by year of menarche), Table 2 (prevalence of depression), and Table 3 (prevalence of depression in girls according to age at menarche), comparing AAM at 11 years and lower to AAM at age 15, assessed at age 15. 
```{r}
# derive numbers in those with AAM 11yrs or less
n10less <- 19196*0.031
n11 <- 19196*0.157

n11less <- n10less+n11

ndep11less <- n11*0.156 + n10less*.206

# derive numbers in reference group
nref <- 19196*0.021
ndepref <- nref*0.089

# odds of depression in those with AAM 11yrs or less
probdep11less <- ndep11less/n11less
probNdep11less <- 1-probdep11less

odds <- probdep11less / probNdep11less

oddsref <- 0.089/(1-0.089)

or <- odds/oddsref

# calculate SE
se <- sqrt(1/ndep11less+ 
             1/(n11less-ndep11less)+
             1/ndepref+
             1/(nref-ndepref))

# calculate CIs
lci = or - 1.96*se
uci = or + 1.96*se

# extract data derived based on the above calculations
OR <- or
OR_low_CI <- lci
OR_up_CI <- uci
log_OR <- log(OR) # convert odds ratios to log odds
log_SE <- ( log(OR_up_CI) - log(OR_low_CI) ) / 3.92 # convert odds ratio SE to log odds SE
var <- log_SE^2 # derive sampling variance 
id <- "kaltiala"
N1 <- 2567 # depression based on table 2
N2 <- 16529 # no depression based on table 2

# transform log odds ratios to standardized beta/correlation coefficients
conversion <- lores(log_OR, var, N1, N2, dig=20, verbose=FALSE)
B <- conversion$r
B_var <- conversion$var.r

# put extracted data into dataframe
kaltiala <- data.frame(B, B_var, id)
kaltiala
```

## Merge effect sizes
```{r}
all_data <- list(joinson, black_klein, stice, lien, kaltiala)
combined <- Reduce(function(...) merge(..., all=T), all_data)
combined
```

## Perform meta-analysis
```{r}
res <- rma.mv(B, B_var, data = combined)
res
```

## Convert effect size to *OR*
First, convert meta-analytic effect size to odds ratio (*OR*) using a function based on equations presented [here](https://www.meta-analysis.com/downloads/Meta-analysis%20Converting%20among%20effect%20sizes.pdf).
```{r}
beta_to_or <- function(beta) {
  d <- (2*beta) / sqrt ( 1 - (beta^2) )
  log_odds <- d * (3.14159 / sqrt(3)) 
  or <- exp(log_odds)
  return(or)
}

beta_to_or(c(res$beta, res$ci.lb, res$ci.ub))
```

## Convert effect size to Cohen's *D*
Then convert effect size from *OR* to *D*.
```{r}
oddsratio_to_d(c(1.657245, 1.511574, 1.818713))
```

## Decide equivalence bounds
**This gives a meta-analytic estimate of *D* = 0.28(0.23-0.33).** According to best practice recommendations, the smallest effect size of interest is determined based on the lower confidence interval of this meta-analytic estimate (*D* = 0.23). The equivalence bounds are based on this value. 

**The lower and upper equivalence bounds are therefore Cohen's *D* -0.23 - 0.23.**